Transitivity of Subtyping for Intersection Types

### 2020-05-14 papers/LMCS-subtyping-intersections/response.txt

>> The abstract mentions that the transitivity proof requires a subtle
>> choice of the induction measure. Nevertheless it seems possible to
>> simplify it by using a lexicographic ordering on (depth B, size B,
>> size C) or on (depth B, size B + size C).

Excellent, we've switched to the lexicographic ordering on
(depth B, size B, size C).

>> page 1, line -5: type systems -> type system
>> page 2, line -15: then then -> then
>> page 3, line 4: provide a link to the Agda development
>> page 3, line 11: functions -> partial functions
>> page 3, lines 18,19: point to Proposition 3(5)
>> page 3, line -7: remarks about eta-reduction are not very clear
>>                  BCD typing is not preserved under eta-reduction
>> page 4, figure 3, A\subseteq B: missing \forall
>> page 5, line -3: <: -> \subseteq (3 times)
>> page 7, lines -2,-1: remove depth(B2)<=depth(B) and

The revision fixes all of the above except for the last item on page 3
regarding eta-reduction. The reviewer is mistaken. BCD typing is
indeed preserved under eta reduction (see Proposition 12.1.15 and
Corollary 14.2.9 of Lambda Calculus with Types by Barendregt, Dekkers,
and Statman). Perhaps the reviewer was thinking of eta-expansion,
which is not preserved by BCD typing.

@@

>> @article{DBLP:journals/iandc/Tiuryn01, author = {Jerzy Tiuryn}, title = {A Sequent Calculus for Subtyping Polymorphic Types}, journal = {Inf. Comput.}, volume = {164}, number = {2}, pages = {345--369}, year = {2001}}

>> Other comments:

>> A. page 1: you say that BCD extends STLC with addition of intersection
>>    types, a top type, and "an infinite collection of type constants".
>>    The last of these surprised me as STLC usually includes base types
>>    (at least one, or else there are no types at all).  If you can
>>    clarify (very briefly -- do not want to distract from the main flow
>>    of your discourse) that would be great.

I've rephrased to remove mention of the STLC, as that was not necessary.

@@

>> B. page 2, you write that the "BCD subtyping judgment does not enjoy
>>    the subformula property".  But proof-theoretically, the subformula
>>    property is a property of a proof system: a proof system has the
>>    subformula property if for every closed proof of that proof system,
>>    every formula occurring in the proof is a subformula of the theorem
>>    proved by the proof.  So it is a (meta-level) type error to refer
>>    to a judgment as having the subformula.  To be clear: it is not
>>    even correct to say that the set of valid subtyping judgments has
>>    the property, for different proof systems (some with, some without
>>    the property) could give rise to the same valid judgments.  We are
>>    speaking about whether some particular set of proof rules has the
>>    property.  Even there we could imagine a distinction (which I did
>>    not find in a brief literature search) between the *rules* having
>>    the property (we could say intrinsic subformula property?) and the
>>    proofs derivable from the rules having the property (extrinsic, the
>>    definition I use just above). For say that the proof system
>>    includes a spurious rule with some premise that is not a subformula
>>    of the conclusion, but where that rule can never actually be
>>    successfully used (due to other constraints in it) in a closed
>>    derivation.  Then the proof system would have the subformula
>>    property extrinsically but not intrinsically.

The mistaken uses of "judgment" have been replaced with "subtyping system"
(i.e. proof system).

@@

>> C. page 3: you say that A \in B "indicates whether A is syntactically
>>    a part of B".  This is not correct without some violence to the
>>    usual notion of syntactic part.  For example, alpha is not \in
>>    (alpha -> alpha) according to the rules of Figure 3.  This is again
>>    part of what I find displeasing about this approach: we have some
>>    very particular and not too intuitive (at least, not correctly
>>    intuitively described, in this case!) notions that yes, do lead to
>>    a correct system (this is proven in Agda).  But it would be much
>>    more satisfying, in the spirit of Tiuryn's work, for example, and
>>    the other works he cites and builds on, to see a compositional
>>    treatment of function types via substructural typing.  This would
>>    avoid these fiddly syntactic definitions.

I've changed the notation for \in and \subseteq to make it clear
that these notions are different from the usual ones.

### 2020-05-15 papers/LMCS-subtyping-intersections/response.txt

>> The introduction is rather short and only gives one application for
>> which the subformula would be interesting. Moreover it is not
>> explained why the proposal from Laurent would not be appropriate for
>> this application.

The revision includes a new section "Motivations from Verified
Compilation" that goes into much more detail regarding the need for
what the revision now calls the "subformula conjunction property", and
explains how we initially used Laurent's proposal but then later
switched to the new system.

>> While the inference system defining <: is claimed to satisfy the
>> subformula property, it is not true strictly speaking. This is because
>> dom(A) and cod(A) are not subformulas of A. E.g. the derivation of
>> (A->B)/\(A->C)<:A->(B/\C) uses A/\A. Most of the consequences of the
>> subformula property hold anyway because dom(A) and cod(A) are not so
>> far from being subformulas of A (they are strictly smaller than A,
>> contain no new atom or new connective, etc).

The revision fixes this problem by being more precise regarding
terminology, defining the "subformula conjunction property".  Also the
revision explains why this property is enough for our purposes.

@@

>> The abstract mentions that the transitivity proof requires a subtle
>> choice of the induction measure. Nevertheless it seems possible to
>> simplify it by using a lexicographic ordering on (depth B, size B,
>> size C) or on (depth B, size B + size C).

Excellent, we've switched to the lexicographic ordering on
(depth B, size B, size C).

>> page 1, line -5: type systems -> type system
>> page 2, line -15: then then -> then
>> page 3, line 4: provide a link to the Agda development
>> page 3, line 11: functions -> partial functions
>> page 3, lines 18,19: point to Proposition 3(5)
>> page 3, line -7: remarks about eta-reduction are not very clear
>>                  BCD typing is not preserved under eta-reduction
>> page 4, figure 3, A\subseteq B: missing \forall
>> page 5, line -3: <: -> \subseteq (3 times)
>> page 7, lines -2,-1: remove depth(B2)<=depth(B) and

The revision addresses all of the above except for the last item on
page 3 regarding eta-reduction. BCD typing is indeed preserved under
eta reduction (see Proposition 12.1.15 and Corollary 14.2.9 of Lambda
Calculus with Types by Barendregt, Dekkers, and Statman). Perhaps the
reviewer was thinking of eta-expansion, which is not preserved by BCD
typing.

@@

Reviewer 2
==========

>> 1. lack of details for the motivating connection to Siek's broader
>>    project(s), alluded to in the Introduction but too briefly to
>>    assess the value of this work on that basis.

The revision includes a new section "Motivations from Verified
Compilation" to address this problem.

>> 2. failure, as also in the Laurent paper that seems to have been the
>>    starting point for this work, to consider the deep line of work on
>>    logical treatments of subtyping.
>>
>> ...  In contrast, Siek's system handles function types through some
>> very particular meta-level operations on type expressions and side
>> conditions to the function subtyping rule (->').  I worry that
>> these conditions will be brittle, and not work well in a setting
>> where compositionality at the type level is required.  Imagine, for
>> example, allowing type-level functions as in System F-omega.  Then
>> particular checks on the forms of type expressions will fail in the
>> presence of applications of type variables (i.e., abstracted type
>> expressions), since we will not have a concrete type expression to
>> apply these various checks and meta-level operations to.  So an
>> approach that is based just on syntactic conditions on type
>> expressions seems fragile and not likely to be extensible to richer
>> type systems.

The logical treatments of subtyping are indeed useful for extending to
richer type systems for programming languages. The author has made use
of them in the context of research on gradual typing.  However, the
motivation for the present work is rather different: constructing
filter models (denotational semantics) for programming languages. For
that purpose, there is no need for richer type systems such as System
F-omega; intersection types already provide what is needed to express
the runtime behavior of functional programs. The revision includes
changes to the introduction to differentiate the varied uses of
intersection type systems and to make it clear that the present
article's motivation comes from denotational semantics, not type
systems for programming languages.

@@

>> B. page 2, you write that the "BCD subtyping judgment does not enjoy
>>    the subformula property".  But proof-theoretically, the subformula
>>    property is a property of a proof system: a proof system has the
>>    subformula property if for every closed proof of that proof system,
>>    every formula occurring in the proof is a subformula of the theorem
>>    proved by the proof.

The mistaken uses of "judgment" have been replaced with "subtyping system"
(i.e. proof system).

>> C. page 3: you say that A \in B "indicates whether A is syntactically
>>    a part of B".  This is not correct without some violence to the
>>    usual notion of syntactic part.  For example, alpha is not \in
>>    (alpha -> alpha) according to the rules of Figure 3.  This is again
>>    part of what I find displeasing about this approach: we have some
>>    very particular and not too intuitive (at least, not correctly
>>    intuitively described, in this case!) notions that yes, do lead to
>>    a correct system (this is proven in Agda).  But it would be much
>>    more satisfying, in the spirit of Tiuryn's work, for example, and
>>    the other works he cites and builds on, to see a compositional
>>    treatment of function types via substructural typing.  This would
>>    avoid these fiddly syntactic definitions.

I've changed the notation for \in and \subseteq to make it clear that
these notions are different from the usual ones and added some
explanation to provide an intuition for what they accomplish.

### 2020-05-15 papers/LMCS-subtyping-intersections/paper.tex
additions to response

\begin{abstract}
  The subtyping rules for intersection types traditionally employ a
  transitivity rule (Barendregt et al. 1983), which means that
  subtyping does not satisfy the subformula property, making it more
  difficult to use in filter models for compiler verification.
  Laurent develops a sequent-style subtyping system, without
  transitivity, and proves transitivity via a sequence of six lemmas
  that culminate in cut-elimination (2018). This article develops a
  subtyping system in regular style that omits transitivity and
  provides a direct proof of transitivity, significantly reducing the
  length of the proof. The new system satisfies something like the
  subformula property: every type occuring in the derivation of $A <:
  B$ is a subformula of $A$ or $B$, or an intersection of such
  subformulas.  Borrowing from Laurent's system, the rule for function
  types is essentially the $\beta$-soundness property. The article
  proves that the new subtyping system is equivalent to that of
  Barendregt, Coppo, and Dezani-Ciancaglini.
\end{abstract}

@@

Intersection types were invented by Coppo, Dezani-Ciancaglini, and
Salle to study normalization in the lambda
calculus~\citep{Coppo:1979aa}. Subsequently intersection type systems
have been used for at least three purposes:
\begin{itemize}
  \item as type systems for programming
    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
  \item to formulate precise static
    analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
    and
  \item to express filter models for the denotational semantics of a
    variety of lambda
    calculi~\citep{Coppo:1980ab,Coppo:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa,Barendregt:2013aa}.
\end{itemize}

@@

\section{Motivations from Verified Compilation}

@@

Our new subtyping judgment relies on a few definitions that are given
in Figure~\ref{fig:aux}. These include the partial functions $\dom{A}$
and $\cod{A}$, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
relations $A \inside B$ and $A \containedin B$.
%
The $\dom{A}$ and $\cod{A}$ partial functions return the domain or
codomain if $A$ is a function type, respectively. If $A$ is an
intersection $A_1 \cap A_2$, then $\dom{A}$ is the intersection of the
domain of $A_1$ and $A_2$.  If $A$ is an atom, $\dom{A}$ is
undefined. Likewise for $\cod{A}$. For example, if $A = (A_1 \to B_1)
\cap \cdots \cap (A_n \to B_n)$, then $\dom{A} = A_1 \cap \cdots \cap
A_n$ and $\cod{A} = B_1 \cap \cdots \cap B_n$.  When $\dom{A}$ or
$\cod{A}$ appears in lemma or theorem statement, we implicitly assume
that $A$ is a type such that $\dom{A}$ and $\cod{A}$ are defined.
%
The $\topP{A}$ predicate identifies types that are equivalent to
$\TOP$. (See Propostion~\ref{prop:top}(\ref{prop:AllBot-⊑-any}).) The
$\topInCod{A}$ predicate identifies types that have $\TOP$ in their
codomain.
%
The relations $A \inside B$ and $A \containedin B$ enable the treatment
of a sequence of intersections $B_1 \cap \cdots \cap B_n$ (modulo
associativity and commutativity) as if it was set of types $\{ B_1,
\ldots, B_n \}$, where each $B_i$ is an atom or a function type.  We
say that $A$ is a part of $B$ if $A \inside B$ and we say that $B$
\emph{contains} $A$ if $A \containedin B$.

>>>>

\begin{abstract}
  The subtyping rules for intersection types traditionally employ a
  transitivity rule (Barendregt et al. 1983), which means that
  subtyping does not satisfy the subformula property, making it more
  difficult to use in filter models for compiler verification.
  Laurent develops a sequent-style subtyping system, without
  transitivity, and proves transitivity via a sequence of six lemmas
  that culminate in cut-elimination (2018). This article develops a
  subtyping system in regular style that omits transitivity and
  provides a direct proof of transitivity, significantly reducing the
  length of the proof. Inspired by Laurent's system, the rule for
  function types is essentially the $\beta$-soundness property.  The
  new system satisfies the ``subformula conjunction property'': every
  type occuring in the derivation of $A <: B$ is a subformula of $A$
  or $B$, or an intersection of such subformulas. The article proves
  that the new subtyping system is equivalent to that of Barendregt,
  Coppo, and Dezani-Ciancaglini.
\end{abstract}

@@

Intersection types were invented by Coppo, Dezani-Ciancaglini, and
Salle to study normalization in the lambda
calculus~\citep{Coppo:1979aa}. Subsequently intersection type systems
have been used for at least three purposes:
\begin{itemize}
  \item as type systems for programming
    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
  \item to formulate precise static
    analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
    and
  \item to express filter models for the denotational semantics of a
    variety of lambda
    calculi~\citep{Coppo:1980ab,Coppo:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa,Barendregt:2013aa}.
\end{itemize}
The motivation for this article comes from the third use of
intersection types, in denotational semantics.  We are interested in
constructing filter models for functional programming languages in
support of mechanized compiler
verification. Section~\ref{sec:motivation} describes how this line of
research motivated the results in this article.

@@

\section{Motivations from Verified Compilation}
\label{sec:motivation}

@@

Our new subtyping judgment relies on a few definitions that are given
in Figure~\ref{fig:aux}. These include the partial functions $\dom{A}$
and $\cod{A}$, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
relations $A \inside B$ and $A \containedin B$.
%
The $\dom{A}$ and $\cod{A}$ partial functions return the domain or
codomain if $A$ is a function type, respectively. If $A$ is an
intersection $A_1 \cap A_2$, then $\dom{A}$ is the intersection of the
domain of $A_1$ and $A_2$.  If $A$ is an atom, $\dom{A}$ is
undefined. Likewise for $\cod{A}$. For example, if $A = (A_1 \to B_1)
\cap \cdots \cap (A_n \to B_n)$, then $\dom{A} = A_1 \cap \cdots \cap
A_n$ and $\cod{A} = B_1 \cap \cdots \cap B_n$.  When $\dom{A}$ or
$\cod{A}$ appears in lemma or theorem statement, we implicitly assume
that $A$ is a type such that $\dom{A}$ and $\cod{A}$ are defined.
%
The $\topP{A}$ predicate identifies types that are equivalent to
$\TOP$. (See Propostion~\ref{prop:top}(\ref{prop:AllBot-⊑-any}).) The
$\topInCod{A}$ predicate identifies types that have $\TOP$ in their
codomain.
%
The relations $A \inside B$ and $A \containedin B$ enable the
treatment of a sequence of intersections $B_1 \cap \cdots \cap B_n$
(modulo associativity and commutativity) as if it were a set of types
$\{ B_1, \ldots, B_n \}$, where each $B_i$ is an atom or a function
type.  We say that $A$ is a part of $B$ if $A \inside B$ and we say
that $B$ \emph{contains} $A$ if $A \containedin B$.
